---
title: "Aged Trial Balance Dashboard | Synthetic Data"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
    theme:
      version: 4
      bootswatch: flatly
      primary: "#1B365D"
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(flexdashboard)
  library(dplyr)
  library(tidyr)
  library(arrow)
  library(plotly)
  library(DT)
  library(crosstalk)
})

as_of  <- as.Date("2026-09-30")
atb    <- read_parquet("data/curated/atb_open_accounts.parquet")
trend  <- read_parquet("data/curated/ar_month_end_trend.parquet")
checks <- readRDS("data/curated/checks.rds")

bucket_levels <- c("0 to 30", "31 to 60", "61 to 90", "91 to 120",
                   "121 to 180", "181 to 365", "Over 365")
status_cols <- c("Unbilled" = "#C9A227", "Billed, awaiting payment" = "#1B365D",
                 "Denied" = "#B5473A", "Underpaid" = "#6C8EBF")
atb <- atb |> mutate(aging_bucket = factor(aging_bucket, bucket_levels),
                     status = factor(status, names(status_cols)))

money <- function(x) {
  ifelse(abs(x) >= 1e6, paste0("$", round(x / 1e6, 1), "M"),
         ifelse(abs(x) >= 1e3, paste0("$", round(x / 1e3), "K"), paste0("$", round(x))))
}
pct <- function(x) paste0(round(100 * x), "%")
over_90 <- sum(atb$balance[atb$aging_bucket %in% bucket_levels[4:7]]) / sum(atb$balance)

plot_layout <- function(p, ...) {
  p |> layout(font = list(family = "Helvetica", size = 12),
              legend = list(orientation = "h", y = -0.2),
              margin = list(t = 10, r = 60), ...) |>
    config(displayModeBar = FALSE)
}
```

Overview
=====================================

Row {data-height=120}
-------------------------------------

### Open AR as of `r format(as_of, "%b %d, %Y")`

```{r}
valueBox(money(sum(atb$balance)), icon = "fa-file-invoice-dollar", color = "primary")
```

### Open accounts

```{r}
valueBox(paste0(round(nrow(atb) / 1000, 1), "K"), icon = "fa-folder-open", color = "info")
```

### AR over 90 days from discharge

```{r}
valueBox(pct(over_90), icon = "fa-hourglass-half",
         color = ifelse(over_90 > 0.35, "danger", "warning"))
```

### Share of AR still unbilled

```{r}
valueBox(pct(sum(atb$balance[atb$status == "Unbilled"]) / sum(atb$balance)),
         icon = "fa-pause-circle", color = "warning")
```

Row
-------------------------------------

### Open AR by aging bucket and status

```{r}
atb |>
  group_by(aging_bucket, status) |>
  summarise(balance = sum(balance), accounts = n(), .groups = "drop") |>
  plot_ly(x = ~aging_bucket, y = ~balance, color = ~status, colors = status_cols,
          type = "bar",
          text = ~paste0(status, "<br>", money(balance), " across ", accounts, " accounts"),
          hoverinfo = "text", textposition = "none") |>
  plot_layout(barmode = "stack", xaxis = list(title = "Days since discharge"),
              yaxis = list(title = "", tickprefix = "$"))
```

### Open AR trend, last 12 month ends

```{r}
trend |>
  mutate(share_90 = ar_over_90 / open_ar) |>
  plot_ly(x = ~month_end) |>
  add_bars(y = ~open_ar, name = "Open AR", marker = list(color = "#1B365D"),
           text = ~money(open_ar), hoverinfo = "text", textposition = "none") |>
  add_lines(y = ~share_90, name = "Share over 90 days", yaxis = "y2",
            line = list(color = "#C9A227", width = 3),
            text = ~pct(share_90), hoverinfo = "text") |>
  plot_layout(xaxis = list(title = ""),
              yaxis = list(title = "", tickprefix = "$"),
              yaxis2 = list(overlaying = "y", side = "right", tickformat = ".0%",
                            range = c(0, 0.6), title = ""))
```

Unbilled
=====================================

Row
-------------------------------------

### Unbilled AR by hold reason

```{r}
atb |>
  filter(status == "Unbilled") |>
  group_by(reason) |>
  summarise(balance = sum(balance), accounts = n(), .groups = "drop") |>
  arrange(balance) |>
  mutate(reason = factor(reason, reason)) |>
  plot_ly(y = ~reason, x = ~balance, type = "bar", orientation = "h",
          marker = list(color = "#C9A227"),
          text = ~paste0(money(balance), " across ", accounts, " accounts"),
          hoverinfo = "text", textposition = "none") |>
  plot_layout(xaxis = list(title = "", tickprefix = "$"), yaxis = list(title = ""))
```

### Unbilled AR by facility and hold reason

```{r}
atb |>
  filter(status == "Unbilled", reason != "Standard bill hold") |>
  group_by(facility_name, reason) |>
  summarise(balance = sum(balance), .groups = "drop") |>
  complete(facility_name, reason, fill = list(balance = 0)) |>
  plot_ly(x = ~reason, y = ~facility_name, z = ~balance, type = "heatmap",
          colors = colorRamp(c("#F4F1E6", "#C9A227", "#7A5C00")),
          text = ~money(balance), hoverinfo = "text+x+y") |>
  colorbar(title = "") |>
  plot_layout(xaxis = list(title = ""), yaxis = list(title = "", autorange = "reversed"))
```

Denials and underpayments
=====================================

Row
-------------------------------------

### Open denials by reason

```{r}
atb |>
  filter(status == "Denied") |>
  group_by(reason) |>
  summarise(balance = sum(balance), accounts = n(), .groups = "drop") |>
  arrange(desc(balance)) |>
  mutate(reason = factor(reason, reason), cum = cumsum(balance) / sum(balance)) |>
  plot_ly(x = ~reason) |>
  add_bars(y = ~balance, name = "Balance", marker = list(color = "#B5473A"),
           text = ~paste0(money(balance), " across ", accounts, " accounts"),
           hoverinfo = "text", textposition = "none") |>
  add_lines(y = ~cum, name = "Cumulative share", yaxis = "y2",
            line = list(color = "#1B365D"), text = ~pct(cum), hoverinfo = "text") |>
  plot_layout(xaxis = list(title = ""), yaxis = list(title = "", tickprefix = "$"),
              yaxis2 = list(overlaying = "y", side = "right", tickformat = ".0%",
                            range = c(0, 1.05), title = ""))
```

### Denied and underpaid AR by payer

```{r}
atb |>
  filter(status %in% c("Denied", "Underpaid")) |>
  group_by(payer_name, status) |>
  summarise(balance = sum(balance), .groups = "drop") |>
  group_by(payer_name) |> mutate(total = sum(balance)) |> ungroup() |>
  arrange(total) |>
  mutate(payer_name = factor(payer_name, unique(payer_name))) |>
  plot_ly(y = ~payer_name, x = ~balance, color = ~status, colors = status_cols,
          type = "bar", orientation = "h",
          text = ~paste0(status, ": ", money(balance)), hoverinfo = "text",
          textposition = "none") |>
  plot_layout(barmode = "stack", xaxis = list(title = "", tickprefix = "$"),
              yaxis = list(title = ""))
```

Account drill down
=====================================

```{r}
detail <- atb |>
  transmute(Account = account_id, Facility = facility_name, Payer = payer_name,
            `Financial class` = financial_class, Status = as.character(status),
            Reason = reason, `Next step` = next_step,
            Discharged = discharge_date, `Aging bucket` = as.character(aging_bucket),
            `Days out` = days_since_discharge, Expected = expected_payment,
            Paid = payments, Balance = balance)
sd <- SharedData$new(detail)
```

Inputs {.sidebar data-width=260}
-------------------------------------

**Filter the open accounts**, then use the Excel or CSV button to export exactly the rows on screen.

```{r}
filter_select("fac", "Facility", sd, ~Facility)
filter_checkbox("status", "Status", sd, ~Status)
filter_select("bucket", "Aging bucket", sd, ~`Aging bucket`)
filter_select("fc", "Financial class", sd, ~`Financial class`)
filter_select("reason", "Hold or denial reason", sd, ~Reason)
```

Row
-------------------------------------

### Open accounts

```{r}
datatable(sd, rownames = FALSE, extensions = "Buttons",
          options = list(dom = "Bfrtip", buttons = list(
                           list(extend = "excel", title = "ATB account detail", text = "Excel"),
                           list(extend = "csv", text = "CSV")),
                         pageLength = 15, scrollX = TRUE),
          class = "compact stripe") |>
  formatCurrency(c("Expected", "Paid", "Balance"), digits = 0) |>
  formatDate("Discharged", method = "toLocaleDateString")
```

Data checks
=====================================

Row
-------------------------------------

### Pipeline validation

Every render runs these checks first. If any fails, the pipeline stops and no dashboard is published.

```{r}
checks |>
  mutate(Result = ifelse(passed, "Pass", "Fail")) |>
  select(Check = check, Result) |>
  datatable(rownames = FALSE, options = list(dom = "t", pageLength = 20),
            class = "compact") |>
  formatStyle("Result", color = styleEqual(c("Pass", "Fail"), c("#2E7D32", "#B5473A")),
              fontWeight = "bold")
```

### About this dashboard

All data is **synthetic**: generated by `R/generate_synthetic_data.R` with a fixed seed. No real patients, accounts, payers or facilities are represented.

**Pipeline:** messy CSV extracts (text dates, currency strings, inconsistent payer spelling, duplicate transactions) are cleaned in R, written to Parquet, and aggregated with DuckDB SQL into the aged trial balance and a 12 month AR trend rebuilt from the transaction history.

**Aging** is measured from discharge so billed and unbilled balances share one clock.

**Source:** [github.com/danielpatterson1/healthcare-atb-dashboard](https://github.com/danielpatterson1/healthcare-atb-dashboard)
